---
title: "Lab 10.1"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Review how to make barplots, scatterplots, and boxplots using `ggplot`.      
2. Use aesthetics to improve readability of plots.          

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Resources
- [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf)
- [ggplot themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)

## Review
We have already learned the basics of ggplot, but let's review the plot types we have learned thus far. We will use the mammal life history data to practice. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*
```{r message=FALSE, warning=FALSE}
#-999 often used in the place of NA, so treat as NA from outset
life_history <- read_csv("data/mammal_lifehistories_v2.csv", na="-999") %>% 
  clean_names()
```

## Bar Plots
Recall that `geom_bar` is used when you want to count the number of observations in a categorical variable. `geom_col` is used when you have already counted the number of observations (or you have a defined x and y) and want to plot those counts.

Make two bar plots showing the number of observations for each order of mammals using both geom types.  

`geom_bar`
```{r}
life_history %>%
  ggplot(mapping=aes(x=order, fill=order))+
  geom_bar(na.rm=T)+
  coord_flip()+
  labs(x="Order", y="Number of Observations", title="Number of Observations per Order")
```

`geom_col`
```{r}
life_history %>%
  count(order, sort=T)  #get observation count for each order
```

```{r}
life_history %>%
  count(order, sort=T) %>%
  ggplot(mapping=aes(x=order, y=n, fill=order))+
  geom_col(na.rm=T)+
  coord_flip()+
  labs(x="Order", y="Number of Observations", title="Number of Observations per Order")
```

What if we wanted a bar plot of the mean mass for each order?
```{r}
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm=T)) %>%
  arrange(desc(mean_mass))
```

```{r}
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm=T)) %>%
  ggplot(mapping=aes(x=order, y=mean_mass, fill=order))+
  geom_col(na.rm=T)+
  coord_flip()+
  labs(x="Order", y="Mean Mass", title="Mean Mass per Order")
```

There are a few problems here. First, the y-axis is in scientific notation. We can fix this by adjusting the options for the session.
```{r}
options(scipen=999) #cancels scientific notation for the session

life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm=T)) %>%
  ggplot(mapping=aes(x=order, y=mean_mass, fill=order))+
  geom_col(na.rm=T)+
  coord_flip()+
  labs(x="Order", y="Mean Mass", title="Mean Mass per Order")
```

Next, the y-axis is not on a log scale. We can fix this by adding `scale_y_log10()`.
```{r}
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm=T)) %>%
  ggplot(mapping=aes(x=order, y=mean_mass, fill=order))+
  geom_col(na.rm=T)+
  coord_flip()+
  scale_y_log10()+
  labs(x="Order", y="Log Mean Mass", title="Log Mean Mass per Order")
```

Lastly, we can adjust the x-axis labels to make them more readable. We do this using `reorder`.
```{r}
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm=T)) %>%
  ggplot(mapping=aes(x=reorder(order, mean_mass), y=mean_mass, fill=order))+
  geom_col(na.rm=T)+
  coord_flip()+
  scale_y_log10()+
  labs(x="Order", y="Log Mean Mass", title="Log Mean Mass per Order")
```

## Scatterplots
Scatter plots allow for comparisons of two continuous variables. Make a scatterplot that compares gestation time and wean mass.
```{r}
life_history %>%
  ggplot(aes(x=gestation, y=wean_mass))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=T)+
  scale_y_log10()+
  labs(x="Gestation Time", y="Wean Mass (g)", title="Gestation Time vs Wean Mass")
```

## Boxplots
Box plots are used to visualize a range of values. So, on the x-axis we have a categorical variable and the y-axis is the range. Make a box plot that compares mass across orders.
```{r}
life_history %>%
  ggplot(aes(x=order, y=log10(mass), fill=order))+  #another way of scaling
  geom_boxplot(na.rm=T)+
  coord_flip()+
  labs(x="Order", y="Log of Mass", title="Log of Mass by Order")
```

## Aesthetics: Labels
Now that we have practiced scatter plots, bar plots, and box plots we need to learn how to adjust their appearance to suit our needs. Let's start with labeling x and y axes.  

```{r}
names(life_history)
```

Is there a relationship between mass and litter size; i.e. do larger mammals have more offspring?
```{r}
life_history %>%
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=F, na.rm=T)+
  scale_x_log10()
```

The plot looks clean, but it is incomplete. A reader unfamiliar with the data might have a difficult time interpreting the labels. To add custom labels, we use the `labs` command.
```{r}
life_history %>%
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=F, na.rm=T)+
  scale_x_log10()+
  labs(x="Mass (g)", y="Litter Size", title="Mass vs Litter Size")
```

We can adjust the plot further by specifying the size and face of the text. We do this using `theme()`. The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` controls the title position.
```{r}
life_history %>%
  ggplot(aes(x=mass, y=litter_size))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=F, na.rm=T)+
  scale_x_log10()+
  labs(x="Mass (g)", y="Litter Size", title="Mass vs Litter Size")+
  theme(plot.title=element_text(size=rel(1.5), hjust=.5))
```

## That's it! Let's take a break and then move on to part 2!  

-->[Home](https://jmledford3115.github.io/datascibiol/)