---
title: "Homework 11"
author: "Arlette Hall"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
``` 

```{r}
options(scipen = 999)
```

## Data
For this homework, we are going to use the [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.  

Let's focus our attention on data from the past 100 years, between 1925 and 2025.
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("data/gapminder.csv") %>% 
  clean_names() %>% 
  filter(between(year, 1925, 2025))
```

1. Use the function(s) of your choice to get an idea of the overall structure of the data, including its dimensions, column names, classes, etc. 
```{r}
glimpse(gapminder)
```

2. How many countries are represented in this dataset?
```{r}
gapminder %>%
  summarize(n_country=n_distinct(country))
#There are 195 countries in the dataset.
```

3. Which five countries have had the largest absolute population growth between 1925-2025? Show this as a table.
```{r}
gapminder %>%
  #use only data of interest (to avoid NA's)
  filter(year == 1925 | year == 2025) %>%
  group_by(country) %>%
  select(country, year, pop) %>%
  #rearrange data so math can be done
  pivot_wider(names_from="year",
              values_from="pop") %>%
  clean_names() %>%
  #abs_pop_growth = 2025 population - 1925 population
  summarize(abs_pop_growth = x2025-x1925) %>%
  #check for & handle negative values
  mutate(abs_pop_growth = ifelse(abs_pop_growth < 0,
                                 abs_pop_growth*(-1),
                                 abs_pop_growth)) %>%
  #show top 5 abs_pop_growth's
  slice_max(abs_pop_growth, n=5)
```

4. Make a plot that shows population growth over time for the top country you found in question #3.
```{r}
library(paletteer)  #for plot themes

#top country is India
gapminder %>%
  filter((year >= 1925 & year <= 2025) & country=="India") %>%
  ggplot(aes(x=year, y=pop/1000000))+
  geom_line(color="purple3")+
  labs(x="Year",
       y="Population (in millions)",
       title="India: Population Growth, 1925-2025")+
  theme_minimal()+
  theme(plot.title=element_text(hjust=0.5, size=rel(1.3), face="bold"),
        axis.title.x=element_text(face="bold"),
        axis.title.y=element_text(face="bold"))+
  scale_x_continuous(breaks=seq(1925, 2025, by=10))
```

5. Has global life expectancy changed between 1925 and 2025? Show the min, median, mean, and max for all countries (combined) in the dataset.
```{r}
gapminder %>%
  filter(year==1925 | year==2025) %>%
  group_by(year) %>%
  summarize(global_lex_min = min(lex, na.rm=T),
            global_lex_median = median(lex, na.rm=T),
            global_lex_mean = mean(lex, na.rm=T),
            global_lex_max = max(lex, na.rm=T))
#Yes, the global life expectancy has increased considerably between 1925 & 2025.
#The min and median are over double what they used to be. The mean is almost double.
#The max has increased by about 33%.
```

6. Make a plot that shows the distribution of life expectancy for all countries in 1925 and 2025 (try using `geom_density`). What do you notice about the shift in life expectancy over time?
```{r}
gapminder %>%
  filter(year==1925 | year==2025) %>%
  mutate(year=factor(year)) %>%
  ggplot(aes(x=lex, color=year, fill=year))+
  geom_density(alpha=0.3, na.rm=T)+
  labs(x="Life Expectancy (Yrs)", y=NULL,
       title="Distribution of Life Expectancy, 1925 & 2025",
       color="Year", fill="Year")+
  theme_minimal()+
  theme(plot.title=element_text(hjust=0.5, size=rel(1.3), face="bold"),
        axis.title.x=element_text(face="bold"),
        axis.title.y=element_text(face="bold"))+
  scale_x_continuous(breaks=seq(20, 90, by=10))
#Life expectancy shifts massively upwards.
#The bottom of the 2025 plot overlaps with the top of the 1925 plot.
```

7. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer in 2025? Show this as a plot. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>%
  filter(country != "Monaco" & year==2025) %>%
  ggplot(aes(x=gdp/1000, y=lex))+
  geom_point(color="turquoise", na.rm=T)+
  geom_smooth(color="purple", method=lm, se=F)+
  labs(x="GDP per Capita (in thousands)", y="Life Expectancy (Yrs)",
       title="GDP per Capita vs Life Expectancy, 2025")+
  theme_minimal()+
  theme(plot.title=element_text(hjust=0.5, size=rel(1.3), face="bold"),
        axis.title.x=element_text(face="bold"),
        axis.title.y=element_text(face="bold"))+
  scale_x_continuous(breaks=seq(0, 150, by=10))
```

8. Which five countries have had the highest absolute GDP per capita growth over the past 50 years? Show this as a table. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>%
  #use only data of interest (to avoid NA's)
  filter(country != "Monaco" & year %in% c(1975,2025)) %>%
  group_by(country) %>%
  select(country, year, gdp) %>%
  #rearrange data so math can be done
  pivot_wider(names_from="year",
              values_from="gdp") %>%
  clean_names() %>%
  #abs_gdp_growth = 2025 gdp - 1975 gdp
  summarize(abs_gdp_growth = x2025-x1975) %>%
  #check for & handle negative values
  mutate(abs_gdp_growth = ifelse(abs_gdp_growth < 0,
                                 abs_gdp_growth*(-1),
                                 abs_gdp_growth)) %>%
  #show top 5 abs_gdp_growth's
  slice_max(abs_gdp_growth, n=5)
```

9. How does per capita GDP growth compare between these same five countries over the past 50 years? Show this as a plot.
```{r}
gapminder %>%
  #use only data of interest
  filter(year %in% c(1975:2025) 
         & country %in% c("Singapore", "Ireland", "Luxembourg", "Guyana", "Norway")) %>%
  group_by(country) %>%
  select(country, year, gdp) %>%
  #plot the data by country
  ggplot(aes(x=year, y=gdp/1000, color=country))+
  geom_line(na.rm=T)+
  #make plot pretty
  labs(x="Year", y="GDP per Capita (in thousands)",
       title="GDP per Capita, 1975-2025",
       color="Country")+
  theme_minimal()+
  theme(plot.title=element_text(hjust=0.5, size=rel(1.3), face="bold"),
        axis.title.x=element_text(face="bold"),
        axis.title.y=element_text(face="bold"))+
  scale_y_continuous(breaks=seq(0, 160, by=20))
```

10. Do one analysis of your choice that includes a table and plot as outputs.
```{r}
#Has the USA's GDP per capita increased or decreased since I was born? (1990-2025) By how much?
gapminder %>%
  #select only data of interest (avoid NA's)
  filter(country == "USA" & year %in% c(1990,2025)) %>%
  group_by(country) %>%
  select(country, year, gdp) %>%
  #rearrange data so math can be done
  pivot_wider(names_from="year",
              values_from="gdp") %>%
  clean_names() %>%
  #gdp_change_1990to2025 = 2025 gdp - 1990 gdp
  summarize(gdp_change_1990to2025 = x2025-x1990)
```

```{r}
gapminder %>%
  #use only data of interest
  filter(year %in% c(1990:2025) & country=="USA") %>%
  #plot the data
  ggplot(aes(x=year, y=gdp/1000))+
  geom_line(color="blue", na.rm=T)+
  #make plot pretty
  labs(x="Year", y="GDP per Capita (in thousands of USD)",
       title="USA: GDP per Capita, 1990-2025")+
  theme_minimal()+
  theme(plot.title=element_text(hjust=0.5, size=rel(1.3), face="bold"),
        axis.title.x=element_text(face="bold"),
        axis.title.y=element_text(face="bold"))+
  scale_y_continuous(breaks=seq(0, 80, by=10))
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 