---
title: "Homework 8"
author: "Arlette Hall"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("naniar")
options(scipen = 999)
```

## About the Data
For this assignment we are going to work with a data set from the [United Nations Food and Agriculture Organization](https://www.fao.org/fishery/en/collection/capture) on world fisheries. These data were downloaded and cleaned using the `fisheries_clean.Rmd` script.  

Load the data `fisheries_clean.csv` as a new object titled `fisheries_clean`.
```{r message=FALSE, warning=FALSE}
fisheries_clean <- read_csv("data/fisheries_clean.csv")
```

1. Explore the data. What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables, etc.? You may use the functions that you prefer.
```{r}
glimpse(fisheries_clean)
```

2. Convert the following variables to factors: `period`, `continent`, `geo_region`, `country`, `scientific_name`, `common_name`, `taxonomic_code`, and `status`.
```{r}
fisheries_clean <- fisheries_clean %>%
  mutate(across(c(period, continent, geo_region, country, scientific_name, common_name, taxonomic_code, status), as.factor))

glimpse(fisheries_clean) #check work
```

3. Are there any missing values in the data? If so, which variables contain missing values and how many are missing for each variable?
```{r}
#skip this for now per instructor's instruction
```

4. How many countries are represented in the data?
```{r}
fisheries_clean %>%
  summarize(n_countries=n_distinct(country))
#There are 249 countries in the data.
```

5. The variables `common_name` and `taxonomic_code` both refer to species. How many unique species are represented in the data based on each of these variables? Are the numbers the same or different?
```{r}
fisheries_clean %>%
  summarize(n_species=n_distinct(common_name))
#There are 3,390 species when using common_name variable.
```

```{r}
fisheries_clean %>%
  summarize(n_species=n_distinct(taxonomic_code))
#There are 3,722 species when using taxonomic_code variable.
#No, the number of species is not the same for the two variables. Common names results in a smaller number of species. This makes sense as common names can repeat (e.g. "Daddy Long Legs" refers to multiple species.)
```


6. In 2023, what were the top five countries that had the highest overall catch?
```{r}
fisheries_clean %>%
  filter(period==2023) %>%
  group_by(country) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  slice_max(total_catch, n=5)
```

7. In 2023, what were the top 10 most caught species? To keep things simple, assume `common_name` is sufficient to identify species. What does `NEI` stand for in some of the common names? How might this be concerning from a fisheries management perspective?
```{r}
fisheries_clean %>%
  filter(period==2023) %>%
  group_by(common_name) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  slice_max(total_catch, n=10)

#NEI stands for "Not Elsewhere Identified".
#You need to know exactly what kinds of animals are in each group, from a wildlife management perspective. This introduces unknowns about what specifically is in each group that has NEI in its name.
```

8. For the species that was caught the most above (not NEI), which country had the highest catch in 2023?
```{r}
fisheries_clean %>%
  filter(common_name=="Alaska pollock(=Walleye poll.)" & period==2023) %>%
  group_by(country) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  slice_max(total_catch)
```

9. How has fishing of this species changed over the last decade (2013-2023)? Create a  plot showing total catch by year for this species.
```{r}
fisheries_clean %>%
  mutate(period = as.numeric(as.character(period))) %>%
  filter(common_name=="Alaska pollock(=Walleye poll.)" & (period>=2013 & period<=2023)) %>%
  group_by(period) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  ggplot(
    mapping=aes(x=period, y=total_catch/1000))+
      geom_point(na.rm=T)+
      geom_smooth(method=lm, se=T)+
      labs(x="Year", y="Total Catch (10^3 tonnes)", title="Alaska Pollock: Total Catch by Year")
#Fishing has increased over the last decade for Alaska Pollock.
```

10. Perform one exploratory analysis of your choice. Make sure to clearly state the question you are asking before writing any code.
```{r}
#Which geographical region caught the most? Is this because the top country is in the top geographical region? Or are there significantly more countries in the top geographical region? Or both?

fisheries_clean %>%  #top geo_regions
  group_by(geo_region) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  arrange(desc(total_catch))

fisheries_clean %>%  #check which countries are in Eastern Asia
  filter(geo_region=="Eastern Asia") %>%
  distinct(country)

fisheries_clean %>%  #top country
  group_by(country) %>%
  summarize(total_catch=sum(catch, na.rm=T)) %>%
  arrange(desc(total_catch))

fisheries_clean %>%  #check number of countries per region
  group_by(geo_region) %>%
  summarize(n_countries=n_distinct(country)) %>%
  arrange(desc(n_countries))

#The top geographical region is Eastern Asia. The top 2 countries are in Eastern Asia (China, Japan). Eastern Asia has fewer countries than most of the geographical regions. So, the reason for Eastern Asia being number 1 is that China and Japan are the top 2.
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  