---
title: "Homework 13"
author: "Arlette Hall"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
``` 

## Data
For this homework, we will use `amniota` data from: Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109. doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r message=FALSE, warning=FALSE}
amniota <- read_csv("data/amniota.csv") %>% 
  clean_names()
```

1. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.
```{r}
glimpse(amniota)
```

```{r}
summary(amniota)
```

2. Make any necessary replacements in the data such that all NA's appear as "NA". 
```{r}
#replace -999 with NA
amniota_no_na <- amniota %>%
  replace_with_na_all(condition = ~.x %in% c(-999))
```

```{r}
#check for more hidden NAs in numeric variables
amniota_no_na %>%
  select(-c(class, order, family, genus, species, subspecies, common_name)) %>%
  summary()
```

```{r}
#check strange 0's in adult_body_mass_g & no_sex_body_mass_g
amniota_no_na %>%
  filter(adult_body_mass_g <= 0.1 | no_sex_body_mass_g <= 0.1) %>%
  select (adult_body_mass_g, no_sex_body_mass_g)
#Looks good.
```

```{r}
#checking lowest values of birth_or_hatching_weight_g
amniota_no_na %>%
  filter(birth_or_hatching_weight_g < 0.01) %>%
  select (birth_or_hatching_weight_g)
#Looks good
```

```{r}
#checking strange negative values for female_maturity_d
amniota_no_na %>%
  filter(female_maturity_d < 0) %>%
  select (female_maturity_d)
```

```{r}
#replace -30258.71100 with NA
amniota_no_na <- amniota_no_na %>%
  mutate(female_maturity_d=na_if(female_maturity_d, -30258.71100))
```

3. How many total NA's are in the data set? Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.
```{r}
library(naniar)
amniota_no_na %>%
  miss_var_summary() %>%
  summarize(n_miss_total = sum(n_miss))
#There are 528,200 NAs in the data set.

amniota_no_na %>%
  miss_var_summary()
```


4. Double check your replacement using `summary`. Do you see any other potential issues? If so, fix them.
```{r}
#just in case there are hidden issues in character columns
amniota_no_na <- amniota_no_na %>%
  replace_with_na_all(condition = ~.x %in% c(common_na_strings))
```

```{r}
amniota_no_na %>%
  summary()
#No more potential issues found.
```

5. Use `naniar` to visualize the number of NA's in each variable. Which variables have the most NA's?
```{r}
amniota_no_na %>%
  miss_var_summary() %>%
  arrange(desc(n_miss))
```

6. Calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?
```{r}
amniota_no_na %>%
  group_by(class) %>%
  select(class, egg_mass_g) %>%
  miss_var_summary() %>%
  arrange(desc(pct_miss))

#Having 100% NA's for egg_mass_g for mammals makes sense (assuming no platypus or echidnas in the data).
#92% NAs for egg_mass_g of reptiles is a bit of a concern. I'd have to spend time looking at the specific species list, as some don't lay eggs or guard their clutches.
#50.1% for birds is understandable as they all guard their nests & can be out of reach. So yes, they have eggs but good luck getting to them.
#These results tell me NA's can be used to get an idea of how detailed (or not detailed) people were in their data collecting. NAs will sometimes reflect the challenges of collecting certain data.
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 