---
title: "Homework 7"
author: "Arlette Hall"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Data 
For this assignment, we will use data from a study on elephants and the effects of poaching on tusk size.  

Reference: [Chiyo, Patrick I., Vincent Obanda, and David K. Korir. “Illegal tusk harvest and the decline of tusk size in the African elephant.” Ecology and Evolution 5, 22: 5216–5229 (2015)](https://doi.org/10.1002/ece3.1769). Data deposited at [Dryad Digital Repository](https://doi.org/10.5061/dryad.h6t7j).  

**1. Before starting data analysis, read the abstract of the paper to get an idea of the questions being asked. In 2-3 sentences, describe what the study is testing and the variables involved.**  
The study wanted to know if poaching affects elephant tusk size. It looked at tusk size in elephants from 1966-1968 and compared it to those born in the mid-1990s (after severe poaching in the 1970s and 80s). Sex, tusk length, basal tusk circumference, years of collection, shoulder height, and estimated age were all variables of interest.

**2. Load `elephants.csv` and store it as a new object called `elephants`.**  

```{r}
library("tidyverse")

elephants <- read_csv("C:/Users/fluff/Desktop/BIS15L_W26_aehall/lab7/data/elephants.csv")
```

**3. Clean the data by converting variable names to lowercase with no spaces or special characters.**  
```{r}
library("janitor")

elephants <- elephants %>% #clean names
  clean_names()
names(elephants) #check names
```

**4. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**   
```{r}
glimpse(elephants)
```

**5. Use `mutate()` Change the variables `years_of_sample_collection`, `elephant_id`, and `sex` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**  
```{r}
elephants <- elephants %>%
  mutate(across(c(years_of_sample_collection, elephant_id, sex), as.factor))

glimpse(elephants) #check work
```

**6. From which years were data collected? Show the sample periods below.**  
```{r}
levels(elephants$years_of_sample_collection)
```

**7. How many males and females were sampled in this study?**  
```{r}
nfemales <- elephants %>%  #count number of females
  filter(sex=="f") %>%
  nrow()

nmales <- elephants %>%  #count number of males
  filter(sex=="m") %>%
  nrow()

#There are 416 females & 361 males in the study.
```

**8. What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?**  
```{r}
elephants %>%
  filter(sex=="f" & years_of_sample_collection=="1966-68") %>%
  select(estimated_age_years, sex) %>%
  summarize(mean_age_f_1966t68=mean(estimated_age_years, na.rm=T),
            median_age_f_1966t68=median(estimated_age_years, na.rm=T),
            sd_age_f_1966t68=sd(estimated_age_years, na.rm=T),
            total=n())

elephants %>%
  filter(sex=="f" & years_of_sample_collection=="2005-13") %>%
  select(estimated_age_years, sex) %>%
  summarize(mean_age_f_2005t13=mean(estimated_age_years, na.rm=T),
            median_age_f_2005t13=median(estimated_age_years, na.rm=T),
            sd_age_f_2005t13=sd(estimated_age_years, na.rm=T),
            total=n())

elephants %>%
  filter(sex=="m" & years_of_sample_collection=="1966-68") %>%
  select(estimated_age_years, sex) %>%
  summarize(mean_age_m_1966t68=mean(estimated_age_years, na.rm=T),
            median_age_m_1966t68=median(estimated_age_years, na.rm=T),
            sd_age_m_1966t68=sd(estimated_age_years, na.rm=T),
            total=n())

elephants %>%
  filter(sex=="m" & years_of_sample_collection=="2005-13") %>%
  select(estimated_age_years, sex) %>%
  summarize(mean_age_m_2005t13=mean(estimated_age_years, na.rm=T),
            median_age_m_2005t13=median(estimated_age_years, na.rm=T),
            sd_age_m_2005t13=sd(estimated_age_years, na.rm=T),
            total=n())

#The sampling looks pretty even between sexes.
#There are significantly fewer elephants in the 2005-13 sampling.
```

**9. Is age (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
ggplot(data=elephants,
       mapping=aes(x=estimated_age_years, y=tusk_length_in_cm, color=years_of_sample_collection))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=T)+
  labs(x="Estimated Age (yrs)", y="Tusk Length (cm)", title="Estimated Age vs Tusk Length by Collection Period")

#Somewhat, yes.
```

**10. Is shoulder height (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
ggplot(data=elephants,
       mapping=aes(x=shoulder_height_in_cm, y=tusk_length_in_cm, color=years_of_sample_collection))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=T)+
  labs(x="Shoulder Height (cm)", y="Tusk Length (cm)", title="Shoulder Height vs Tusk Length by Collection Period")

#Yes, mostly.
```

**11. The authors argue that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length. Is this supported by the data? Show your code and calculations below.**  
```{r}
elephants1966 <- elephants %>%
  filter(years_of_sample_collection=="1966-68")
elephants2005 <- elephants %>%
  filter(years_of_sample_collection=="2005-13")

elephants1966 %>%
  summarize(mean_tusk_length_1966t68=mean(tusk_length_in_cm, na.rm=T),
            total=n())
elephants2005 %>%
  summarize(mean_tusk_length_2005t13=mean(tusk_length_in_cm, na.rm=T),
            total=n())

#Yes, this is supported by the data.
```

**12. Male elephants reach effective sexual maturity at 25 years while females are sexually mature at 12 years. Make a new dataframe that extracts only the males and females at sexual maturity. Then, make a plot that shows the range of tusk length between the two sample periods for these mature elephants.**   
```{r}
mature_elephants <- elephants %>%
  filter((sex=="f" & estimated_age_years>=12) | (sex=="m" & estimated_age_years>=25))

ggplot(data=mature_elephants,
       mapping=aes(x=estimated_age_years, y=tusk_length_in_cm, color=years_of_sample_collection))+
  geom_point(na.rm=T)+
  geom_smooth(method=lm, se=T)+
  labs(x="Estimated Age (yrs)", y="Tusk Length (cm)", title="Mature Elephants: Estimated Age vs Tusk Length by Collection Period")
```

## Submit the Homework
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the homework assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository. 